---
title: "Build your Table 1 in R"
description: |
  A quick tutorial for clinicians new to R
author:
  - name: Jordan Gauthier
    url: https://github.com/drjgauthier
    affiliation: Fred Hutch/UW
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import dataset
First let's import a dataset. We will use open-access data from the CIBMTR:
*HLA haploidentical versus matched unrelated donor transplants with post-transplant cyclophosphamide based prophylaxis.*
https://www.cibmtr.org/ReferenceCenter/PubList/PubDsDownload/Pages/default.aspx

```{r importpckg}
library(tidyverse)
library(gtsummary)
library(haven)
df <- read_sas("gs1801_v2.sas7bdat",NULL)
```

## Glimpse the data
Two very useful functions: dplyr::glimpse() and skimr::skim()
Let's give it a try!

```{r glimpse,layout="l-body-outset"}
glimpse(df)
```

## Skim the data
We are shown a list of variable names, variable type (here, numeric; "dbl" stands for double class. A double-precision floating point number.) We are also shown the first 15 rows.

```{r skim}
skimr::skim(df)
```

The skim() function is another great tool to explore your dataset. Descriptive statistics are given for each variables, including number of missing values.
Now that we know the dataset is looking nice and clean (particularly variable names!), let's build the table using the great {gtsummary} package.

## Build Table 1 (Patient and transplant characteristics)

```{r,layout="l-body-outset"}
df %>% 
  transmute( # transmute() allows use to both select and transform variables
    Sex=ifelse(sex==1,"Male","Female"), #ifelse() is convenient to recode binary variable
    Age=age,
    `Performance status (Karnosfky Scale)`=ifelse(kps==1,"90-100","<90"), #Back ticks allow you include spaces in variable names
    `Disease type`=recode(disease,"1"="AML","2"="ALL","3"="MDS")
  ) %>% 
    tbl_summary(
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75}) \n {min}-{max}",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) %>% 
  bold_labels() 
```

### Split and compare groups in Table 1

```{r,layout="l-body-outset"}
df %>%
  transmute(
    # transmute() allows use to both select and transform variables
    Sex = ifelse(sex == 1, "Male", "Female"),
    #ifelse() is convenient to recode binary variable
    Age = age,
    `Performance status (Karnosfky Scale)` = ifelse(kps == 1, "90-100", "<90"),
    `Disease type` = recode(
      disease,
      "1" = "AML",
      "2" = "ALL",
      "3" = "MDS"
    ),
    `Donor group` = ifelse(main == 1, "Haplo PTCy/CNI/MMF", "MUD PT-Cy/CNI/MMF")
  ) %>%
  tbl_summary(
    by = `Donor group`,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75}) \n {min}-{max}",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) %>%
  bold_labels() %>%
  add_p() %>%
  add_stat_label()
```

*More resources:*  

{gtsummary} to build/tweak your tables [https://www.danieldsjoberg.com/gtsummary/reference/index.html](https://www.danieldsjoberg.com/gtsummary/reference/index.html)  

{dplyr} for data manipulation [https://dplyr.tidyverse.org/](https://dplyr.tidyverse.org/)  
